\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-04-14 main.tex
notes on sections

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

\title{
\bf  \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\}
\date{}
\author{Christian Hilbe, Nikoleta E. Glynatsi}

@@

Evolutionary game theory is a mathematical framework for modeling evolution in
biological, social and economical systems. %ToDo references
The success of individuals is not constant, but depends on the composition of
the population. Individuals interact according to rules of the game and their
payoff is translated into reproductive fitness.

Traditionally, evolutionary game studies work with idealized assumptions when
calculating fitness. As stated in~\cite{Roca2006}, it is commonly assumed that
individuals play many times and with all other players before reproduction takes
place, so that payoffs, equivalently fitness, are given by the mean distribution
of types in the population. Initially, this implies that selection occurs much
more slowly than the interaction between individuals, even experimental studies
show that this may not always be the case in biology. Based on %ToDo references
it is clear that in cultural evolution or social learning the time scale of
selection is much closer to the time scale of interaction. The work
of~\cite{Roca2006} relaxed this assumption and showed that rapid selection
affects evolutionary dynamics in such a dramatic way that for some games it even
changes the stability of equilibria.

Furthermore, the second assumption implies that individuals have the time to
interact with with all other players. Moreover, this implies that individuals
when update their strategies over time, individuals are assumed to have perfect
memory. However, when modeling how individuals make decisions in each round,
these models assume that players only remember the last
round~\cite{Traulsen2007, Baek2016, Ye2017}. The aim of this work is too relax
this assumption, and to further the work of~\cite{Roca2006} by also considering
strategies that interact to the past interactions.

In literature, when individuals update their strategies over time they interact
with all other players so that the are given by the mean distribution of types
in the population. These payoffs are referred to as the \textit{expected}
payoffs of individuals. We contrast this standard scenario with an alternative
scenario where individuals update their strategies when remembering several of
their past interactions but not all of them. We refer to these as the \textit{stochastic}
payoffs of individuals.

Evolutionary game dynamics have extensively been studied from mostly
deterministic models based on rate equations to stochastic individual-based
models. These more sophisticated models make use of different rules to determine
how successful strategies spread. As stated in~\cite{Wu2015} there are two
classes have been used extensively, \textit{fitness-based processes} in which an
individual chosen proportional to fitness reproduces and the offspring replaces
a randomly chosen individual and \textit{pairwise comparison} processes in which
a pair of individuals is chosen, and where subsequently one of these individuals
may adopt the strategy of the other. This adoption occurs with a probability
that depends on the payoff of both individuals, such that better players are
more likely to be imitated than those who do worse.

In this we work we follow a pairwise comparison processes. We consider a well
mixed population of $N$ individuals. In our model mutation happens rarely and
such each individual can be of one of the two types, either of the
\textit{resident} or either of the \textit{mutant}.

Regarding the game, we the important case of symmetric \(2 \times 2\) games,
in which the payoffs are given by the following matrix:

@@

These cases include the Snowdrift~\cite{sugden2004economics}, Stag Hunt~\cite{skyrms2001stag},
Harmony~\cite{Licht1999} and the Prisoner's Dilemma games. These results are 
presented in Section~\ref{section:2_by_2_games}.

Moreover, we will specifically consider the donation game where each player can
cooperate by providing a benefit \(b\) to the other player at their cost \(c\),
with \(0 < c < b\). Then, \(T=b, R=b-c, S=-c, P=0\), and matrix~(\ref{eq:games})
is give by:

@@

These results are presented in Section~\ref{section:donation}. The relative
influence of the game is controlled by an external parameter, the so called
intensity of selection $\beta$.

The space of strategies in repeated is infinite. To continue with our
evolutionary analysis, we assume herein that individuals at most make use of
simple \textit{reactive strategies}. Reactive strategy are a set of memory-one
strategies that only take into account the previous action of the opponent. An
example of a reactive strategy is Tit For Tat. Reactive strategies can be
written explicitly as a vector \(\in \R_{3}\). More specifically a reactive
strategy \(s\) is given by \(s=(y, p, q)\) where \(y\) is the probability that
the strategy opens with a cooperation and \(p, q\) are the probabilities that
the strategy cooperates given that the opponent cooperated and defected
equivalently. The individuals of our population do not interact for a given
number of turns, on the contrast, we consider that each match has a probability
$\delta$ to continue following each turn.

\section{Expected and stochastic payoffs in the donation game}\label{section:donation}

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulation results
for the above described process. Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving conditional
cooperation probabilities $p$ and $q$ (assuming that the discount
factor~$\delta$ and the benefit $b$ are comparably high). The left panel
corresponds to the standard scenario considered in the previous literature. It
considers players who use expected payoffs to update their
strategies. The right panel shows the scenario considered herein, in which
players update their strategies based on their last round's payoff. The figure
suggests that when updating is based on expected payoffs, players tend to be
more generous (their $q$-values are higher on average). In addition, players are
generally more cooperative.

\begin{figure}[!htbp]
    \includegraphics[width=\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    We have run two simulations of the evolutionary process described in
    Section~\ref{section:methods} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. Also the resulting cooperation rate is smaller. On
    average, players cooperate roughly in half of all rounds. Parameters:
    $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

The difference in the cooperation rates becomes more and more obvious as the
temptation to defect increases, Figure~\ref{fig:expected_and_stochastic_for_donation_b_10}.

\begin{figure}[!htbp]
    \includegraphics[width=\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    Parameters:
    $N\!=\!100$, $b\!=\!10$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation_b_10}
\end{figure}

We explored how the evolving cooperation rates change as we vary the benefit $b$
and the selection strength $\beta$
(Figure~\ref{fig:cooration_under_parameters}). In all cases, we find that the
two scenarios yield similar cooperation rates when the respective parameters are
small. Once there is a high benefit to cooperation, strong selection, or a high
expected number of rounds, updating based on expected payoffs yields higher
cooperation rates.

\begin{figure}
    \centering
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
    \end{subfigure}
\caption{{\bf The evolution of cooperation for different parameter values.} 
While the previous figures depict the evolutionary outcome for fixed parameter
values, here we vary the benefit of cooperation $b$ and the strength of
selection $\beta$. In all cases, stochastic payoff evaluation tends to reduce
the evolving cooperation rates. Unless explicitly varied, the parameters of the
simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$,
$\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$ time steps for
each parameter combination.}\label{fig:cooration_under_parameters}
\end{figure}

@@

\section{Methods}\label{section:methods}

@@

@@

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

\title{
\bf  \sffamily \LARGE Evolution of cooperation among individuals with limited payoff memory\\}
\date{}
\author{Christian Hilbe, Nikoleta E. Glynatsi, Alex McAvoy}

@@

Evolutionary game theory is a mathematical framework for modeling evolution in
biological, social and economical systems. %ToDo references
In evolutionary models, the success of an individual is not constant but it
depends on the composition of the population. Individuals interact according to
the rules of a pre defined game and they payoffs they achieved are translated
into fitness.

Evolutionary models often work with idealized assumptions when estimating the
fitness of a individual. It is commonly assumed that an individual will interact
several times and with all other players in the population. Equivalently their
fitness is given by the mean payoff they achieved over the different types in
the population after multiple intercalations. These payoff is referred to as the
\textit{expected} payoff.

Initially this assumption implies that selection occurs much more slowly than
the interaction between individuals. However, experimental studies show that
this may not always be the case in biology %ToDo references.
In~\cite{Roca2006} this assumption is realized and it shows that rapid selection
affects evolutionary dynamics in such a dramatic way that for some games it even
changes the stability of equilibria.

Moreover, estimating expected payoffs assume that individuals have a perfect a
memory. Their recall several of their interactions with each player in the
population. This creates a curious inconsistency. With a only few notable
exceptions~\cite{Hauert1997, Stewart2016}, evolutionary models focus on naive
subjects who can only choose from a restricted set of
strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016} when modeling how individuals
make decisions in each round. However, when modeling how individuals update
their strategies over time, individuals are assumed to have perfect memory.

We will explore how robust our understanding of cooperation is as models deviate
from the perfect memory assumption. We start by considering two extreme
scenarios. The first is the classical scenario of the expected payoffs and the
alternative scenario where individuals update their strategies only based on the
very last payoff they obtained. For our second approach we allow individuals to
use more memory. We refer to these as the \textit{stochastic} payoffs of
individuals.

Early results for the repeated prisoner’s dilemma and for the snowdrift game
suggest that memory size can have a considerable effect on the evolutionary
dynamics. As a rule of thumb, we observe that individuals with limited memory
tend to adopt less generous strategies and they achieve less cooperation than in
the classical scenario. In contrast, in the stag-hunt and the harmony game, the
impact of memory is less striking. Here individuals with limited memory perform
nearly as well as individuals with full memory.

\section{Model Setup}

Evolutionary game dynamics have extensively been studied using stochastic
individual-based models that make use of different rules to determine how
successful strategies spread. There are two classes have been used extensively
based on~\cite{Wu2015} (i) \textit{fitness-based processes} in which an
individual chosen proportional to fitness reproduces and the offspring replaces
a randomly chosen individual (ii) \textit{pairwise comparison} processes in
which a pair of individuals is chosen, and where subsequently one of these
individuals may adopt the strategy of the other. This adoption occurs with a
probability that depends on the fitness of both individuals, such that better
players are more likely to be imitated than those who do worse.

We us a pairwise comparison processes with a population of $N$ individuals. In
our model mutation happens rarely and such each individual can be of one of the
two types, either of the \textit{resident} or either of the \textit{mutant}.

The fitness of the individuals will be based on the payoffs they receive from
their interactions. he individuals of our population do not interact for a given
number of turns, on the contrast, we consider that each match has a probability
$\delta$ to continue following each turn. In each turn individuals can decide whether
to cooperate (C) or to defect (D). The
payoffs in a given round depend on the player's and co players decisions. We
consider all the important case of symmetric \(2 \times 2\) games, in which the
payoffs are given by the following matrix:

@@

These cases include the Snowdrift~\cite{sugden2004economics} ($T>R>P>S$), Stag
Hunt~\cite{skyrms2001stag} ($T>R>P>S$), Harmony~\cite{Licht1999} ($T>R>P>S$) and
the Prisoner's Dilemma ($T>R>P>S$) games. The results based on these well know
\(2 \times 2\) games are discussed in Section~\ref{section:2_by_2_games}.

In Section~\ref{section:donation}, we will specifically consider the donation
game where each player can cooperate by providing a benefit \(b\) to the other
player at their cost \(c\), with \(0 < c < b\). Then, \(T=b, R=b-c, S=-c, P=0\),
and matrix~(\ref{eq:game}) is give by:

@@

The relative influence of the payoffs on the adopt the strategy of the other is
controlled by an external parameter, the so called intensity of selection
$\beta$.

The space of strategies in repeated is infinite. To continue with our
evolutionary analysis, we assume herein that individuals at most make use of
simple \textit{reactive strategies}. Reactive strategy are a set of memory-one
strategies that only take into account the previous action of the opponent.
Reactive strategies can be written explicitly as a vector \(\in \R_{3}\). More
specifically a reactive strategy \(s\) is given by \(s=(y, p, q)\) where \(y\)
is the probability that the strategy opens with a cooperation and \(p, q\) are
the probabilities that the strategy cooperates given that the opponent
cooperated and defected equivalently.

\section{Analysis of Stochastic Payoffs in the Donation game}\label{section:donation}

We first explore the effect of stochastic payoffs on the cooperative behaviour
by considering the extreme case that an individual's fitness is based on their
last interaction with one other individual. We compare this to the expected
payoffs. Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulation
results for the described process of Section~\ref{section:methods}.

Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$ (assuming that the discount
factor~$\delta$ and the benefit $b$ are comparably high). The left panel
corresponds to the standard scenario considered in the previous literature. It
considers players who use expected payoffs to update their strategies. The right
panel shows the scenario considered herein, in which players update their
strategies based on their last round's payoff. The figure suggests that when
updating is based on expected payoffs, players tend to be more generous (their
$q$-values are higher on average). In addition, players are generally more
cooperative.

There is a statistically significant different in the cooperation distribution of
the population over time. The cooperation rate falls significantly yo under 50\%.
There results are true for a strong selection strength ($\beta=10$). The follow
up question that arises is what effect does the selection strength have on the
cooperation rate?

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    We have run two simulations of the evolutionary process described in
    Section~\ref{section:methods} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. Also the resulting cooperation rate is smaller. On
    average, players cooperate roughly in half of all rounds. Parameters:
    $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

Figure~\ref{fig:cooperation_rate_over_betas} illustrates the results for
different runs of the evolutionary process where $\beta\in\{10 ^ -2, 10 ^ - 1, 10
^ 0, 10 ^ 1, 10 ^ 2\}$. In the case where $\beta=10 ^ -2$ the process is almost
random, as the effect of the payoffs is very small. The cooperation rate for
this case and the case of $\beta=10 ^ -1$ between the stochastic and expected
payoffs are the same. However, it is clear that this does not apply given 
$\beta \geq 1$. The difference in the cooperation rate increases, in the case
where for the stochastic payoffs it can drop to 10\%.

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
      \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
    \end{subfigure}
\caption{{\bf The evolution of cooperation for different parameter values.} 
While the previous figures depict the evolutionary outcome for fixed parameter
values, here we vary the benefit of cooperation $b$ and the strength of
selection $\beta$. In all cases, stochastic payoff evaluation tends to reduce
the evolving cooperation rates. Unless explicitly varied, the parameters of the
simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$,
$\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$ time steps for
each parameter combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

Figure~\ref{fig:cooperation_rate_over_betas} illustrates the results for
different runs of the evolutionary process where $\beta\in\{10 ^ -2, 10 ^ - 1, 10
^ 0, 10 ^ 1, 10 ^ 2\}$. In the case where $\beta=10 ^ -2$ the process is almost
random, as the effect of the payoffs is very small. The cooperation rate for
this case and the case of $\beta=10 ^ -1$ between the stochastic and expected
payoffs are the same. However, it is clear that this does not apply given 
$\beta \geq 1$. The difference in the cooperation rate increases, in the case
where for the stochastic payoffs it can drop to 10\%.

\begin{figure}[!htbp]
    \centering
    \begin{subfigure}{.65\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
      \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
    \end{subfigure}%
    \begin{subfigure}{.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
    \end{subfigure}
\caption{{\bf The evolution of cooperation for different parameter values.} 
While the previous figures depict the evolutionary outcome for fixed parameter
values, here we vary the benefit of cooperation $b$ and the strength of
selection $\beta$. In all cases, stochastic payoff evaluation tends to reduce
the evolving cooperation rates. Unless explicitly varied, the parameters of the
simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$,
$\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$ time steps for
each parameter combination.}
\end{figure}

@@

\appendix

\section{Model Setup}\label{section:methods}

@@

\section{Expected Payoffs}

@@

\section{Stochastic Payoffs}
